---
title: "GeneFunnel: scientific scoring and practical use"
author: "Emir Turkes"
date: "18 July 2026"
package: genefunnel
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
    math_method: mathml
vignette: >
  %\VignetteIndexEntry{GeneFunnel: scientific scoring and practical use}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Assisted-by: OpenAI Codex. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.align = "center"
)
Sys.setenv(TZ = "UTC")
```

# Introduction

GeneFunnel turns a non-negative feature-by-sample matrix into a
gene-set-by-sample score matrix. It preserves observed magnitude while
penalizing absolute variation among the available members of each set. Every
gene set and sample is calculated independently.

Transcriptomic and proteomic assays commonly use Bioconductor's matrix,
sparse-matrix, and parallel infrastructure. GeneFunnel fits that environment by
accepting base and `Matrix` matrices and a `BiocParallelParam`, while returning
a plain numeric matrix for downstream analysis. It is a scorer only: it does
not normalize data, impute missing values, map identifiers, retrieve gene sets,
or perform statistical testing.

The package is in development and has not been accepted into Bioconductor. Its
source and issue tracker are on the
[GeneFunnel repository](https://github.com/eturkes/genefunnel).

# Installation

Install the development version from GitHub with Bioconductor-compatible
dependencies:

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("eturkes/genefunnel")
```

Load the installed package. Choose a serial backend for the first examples.

```{r load}
library(genefunnel)
serial <- BiocParallel::SerialParam()
```

# Score definition

For one declared gene set, GeneFunnel first deduplicates member identifiers and
matches them exactly and case-sensitively to matrix row names. Within sample
$j$, it omits matched values that are `NA` or `NaN`. If the remaining observed
values are $x_1,\ldots,x_n$ and $n \ge 2$, the score is

$$
\bar{x}=\frac{1}{n}\sum_{i=1}^{n}x_i
$$

and

$$
\operatorname{GeneFunnel}(x_1,\ldots,x_n)
=\sum_{i=1}^{n}x_i
-\frac{n}{2(n-1)}\sum_{i=1}^{n}|x_i-\bar{x}|.
$$

The cell is `NA_real_` when fewer than two observed members remain. Zero is an
observed value, including an implicit sparse zero, so it participates in $n$,
the sum, the mean, and the deviation. Infinities and negative values are
invalid.

The following matrix places the canonical cases in columns.

```{r canonical}
canonical <- matrix(
    c(4, 4, 4, 0, 0, 0, 4, 0, 0, 1, 2, NA, 4, NA, NA),
    nrow = 3,
    dimnames = list(
        c("A", "B", "C"),
        c("equal", "zeros", "one_nonzero", "missing_pair", "one_observed")
    )
)
canonical_sets <- list(
    three_members = c("A", "B", "C"),
    two_members = c("A", "B")
)

canonical_scores <- genefunnel(
    canonical,
    canonical_sets,
    BPPARAM = serial
)
knitr::kable(canonical_scores, digits = 3)

stopifnot(
    identical(
        unname(canonical_scores["three_members", ]),
        c(12, 0, 0, 2, NA_real_)
    ),
    identical(
        unname(canonical_scores["two_members", ]),
        c(8, 0, 0, 2, NA_real_)
    )
)
```

Equal values score their sum. All-zero sets and sets with only one non-zero
member score zero. The missing-pair column scores the observed values `c(1, 2)`
as 2; the last column has only one observed value and returns `NA`.

# Explain score components

`genefunnel_components()` calculates the same authoritative score while
exposing the arithmetic facts hidden by one scalar. The diagnostic native path
reads each input chunk once and returns aligned matrices.

```{r components}
canonical_components <- genefunnel_components(
    canonical,
    canonical_sets,
    BPPARAM = serial
)
component_table <- data.frame(
    sample = colnames(canonical),
    score = canonical_components$score["three_members", ],
    observed_sum = canonical_components$observed_sum["three_members", ],
    penalty = canonical_components$penalty["three_members", ],
    balance = canonical_components$balance["three_members", ],
    effective_size = canonical_components$effective_size["three_members", ],
    observed_fraction = canonical_components$observed_fraction[
        "three_members",
    ],
    row.names = NULL,
    check.names = FALSE
)
knitr::kable(component_table, digits = 3)

stopifnot(
    identical(canonical_components$score, canonical_scores),
    identical(
        unname(canonical_components$observed_sum["three_members", ]),
        c(12, 0, 4, 3, 4)
    ),
    is.na(canonical_components$balance["three_members", "zeros"]),
    identical(
        canonical_components$status$semantic[
            "three_members",
            "one_observed"
        ],
        "too_few_observed"
    )
)
```

For a positive total, `balance` is Bulla's evenness index on the observed
within-set shares and the exact factorization is `score = observed_sum *
balance`. It describes arithmetic evenness, not co-expression or pathway
truth. `observed_fraction` uses globally matched members as its denominator;
declared-set coverage remains a separate fact.

Finite sums, penalties, or balances can overflow or underflow an R double even
when the score is representable. `status` distinguishes ordinary, binary-scaled,
and unavailable diagnostics. A scaled value uses the corresponding `mantissa`
and `exponent` matrices under `canonical_components$scaled`; ordinary values
have `NA` there. The `conditioning` matrix marks cells where rounded component
reconstruction is unsafe.

# Coverage is caller controlled

A declared member absent from the matrix changes global assay coverage. It is
different from a missing cell, which affects only one sample, and from an
observed zero. `gene_set_coverage()` exposes these facts before scoring.

```{r coverage}
catalogue <- list(
    complete = c("A", "B"),
    partial = c("A", "B", "D"),
    duplicated = c("A", "A", "B"),
    insufficient = c("A", "Z")
)
coverage <- gene_set_coverage(catalogue, rownames(canonical))
knitr::kable(coverage, digits = 3)
```

GeneFunnel can score a set whenever at least two unique members match. It does
not impose a coverage fraction because assay-specific policies differ. A strict
transcriptomic analysis might require complete coverage:

```{r strict-coverage}
strict <- !is.na(coverage$coverage) & coverage$coverage == 1
names(catalogue[strict])
```

A low-coverage proteomic analysis might require at least 50% coverage and at
least two matched members:

```{r relaxed-coverage}
relaxed <- !is.na(coverage$coverage) &
    coverage$coverage >= 0.5 &
    coverage$matched_size >= 2
names(catalogue[relaxed])

relaxed_scores <- genefunnel(
    canonical,
    catalogue[relaxed],
    BPPARAM = serial
)
knitr::kable(relaxed_scores, digits = 3)
```

These thresholds are examples, not package recommendations. Record the exact
gene-set collection and version alongside the chosen policy.

# Dense and sparse matrices

Sparse input stays sparse through scoring, and implicit zeros remain observed.
Stored sparse `NA` and `NaN` values follow the same sample-specific omission
rule as dense missing values.

```{r sparse}
sparse_canonical <- Matrix::Matrix(canonical, sparse = TRUE)
sparse_scores <- genefunnel(
    sparse_canonical,
    canonical_sets,
    BPPARAM = serial
)

isTRUE(all.equal(
    sparse_scores,
    canonical_scores,
    tolerance = 0,
    check.attributes = TRUE
))
```

The dense and sparse implementations are required to produce the same named
matrix. Sparse storage changes memory use and traversal, not the scientific
calculation.

# Serial and parallel execution

`BPPARAM` accepts any suitable `BiocParallelParam`. `SerialParam()` is useful
for debugging and small inputs. A portable SOCK backend can distribute bounded
column chunks while preserving sample order.

```{r parallel}
sock <- BiocParallel::bpstart(BiocParallel::SnowParam(
    workers = 2L,
    type = "SOCK",
    progressbar = FALSE
))
parallel_scores <- tryCatch(
    genefunnel(canonical, canonical_sets, BPPARAM = sock),
    finally = BiocParallel::bpstop(sock)
)

identical(parallel_scores, canonical_scores)
```

Process startup and data transfer can outweigh parallel work on small inputs.
Benchmark the intended workload rather than assuming that more workers are
faster.

# Benchmark design and interpretation

The thesis cited with this package includes Sections 4.5--4.9. These sections
progress from controlled toy matrices through one known-truth simulation and
controlled edits on a real background. They continue with an unmodified-data
plausibility audit and resource scaling. The
table turns those performed concepts into recommended evidence layers. Replicate
uncertainty, held-out evaluation, explicit dense/sparse and overlap workloads,
process-aware memory, and output identity below are strengthened requirements,
not claims about the thesis execution. Each layer answers a different question:

```{r benchmark-layers, echo=FALSE}
benchmark_layers <- data.frame(
    Layer = c(
        "Exact properties",
        "Synthetic truth",
        "Controlled real background",
        "Unmodified real data",
        "Computational scaling"
    ),
    Design = c(
        "Equal, zero, missing, and dispersed hand-constructed profiles",
        paste(
            "Planted signal/null sets under partial activation,",
            "mean-preserving variance reduction, and coordinated changes"
        ),
        paste(
            "Known edits to isolated and overlapping sets",
            "in measured data"
        ),
        "Biologically motivated contrasts plus member-level inspection",
        "Repeated storage, backend, size, and overlap workloads"
    ),
    `Primary readout` = c(
        "Formula and ordering assertions",
        "Recovery, false discoveries, and uncertainty across replicates",
        "Target recovery and shared-member spillover",
        "Plausibility and failure anatomy",
        paste(
            "Elapsed time, allocation, process-tree memory,",
            "and output identity"
        )
    ),
    `Evidence boundary` = c(
        "Score semantics only",
        "The declared generator and analysis pipeline",
        "Responsiveness to the edit, not biological truth",
        "Exploratory evidence only",
        "The recorded workload, software, and hardware"
    ),
    check.names = FALSE
)
knitr::kable(benchmark_layers, row.names = FALSE)
```

The layers do not rescue one another. An exact property test cannot establish
biological usefulness. A planted signal cannot establish performance outside
its generator. A plausible real-data result is not ground truth.

## A controlled pattern benchmark

The following small fixture gives arithmetic analogues of the three signal
mechanisms in the thesis's synthetic work. It is not a group-level simulation
or detection test. The reference is balanced. The coordinated profile raises
every member. Partial activation raises only half. The same-sum profile
redistributes a fixed total. At that fixed total, the one-nonzero profile is the
maximal-deviation boundary.

The matched-total pair increases deviation from a balanced reference, whereas
the thesis perturbation reduced intergene variance. The direction differs, but
both isolate score response while the set mean and total stay fixed.

```{r controlled-benchmark}
benchmark_profiles <- cbind(
    reference = c(2, 2, 2, 2),
    coordinated_shift = c(2.25, 2.25, 2.25, 2.25),
    partial_activation = c(3, 3, 2, 2),
    same_sum_uneven = c(6, 1, 1, 0),
    one_nonzero = c(8, 0, 0, 0)
)
rownames(benchmark_profiles) <- paste0(
    "G",
    seq_len(nrow(benchmark_profiles))
)
benchmark_set <- list(pathway = rownames(benchmark_profiles))

pattern_scores <- genefunnel(
    benchmark_profiles,
    benchmark_set,
    BPPARAM = serial
)
pattern_components <- genefunnel_components(
    benchmark_profiles,
    benchmark_set,
    BPPARAM = serial
)
pattern_table <- data.frame(
    scenario = colnames(benchmark_profiles),
    observed_sum = pattern_components$observed_sum["pathway", ],
    penalty = pattern_components$penalty["pathway", ],
    balance = pattern_components$balance["pathway", ],
    score = pattern_components$score["pathway", ],
    row.names = NULL,
    check.names = FALSE
)
knitr::kable(pattern_table, digits = 3)

isolated_scores <- vapply(
    seq_len(ncol(benchmark_profiles)),
    function(j) {
        genefunnel(
            benchmark_profiles[, j, drop = FALSE],
            benchmark_set,
            BPPARAM = serial
        )["pathway", 1L]
    },
    numeric(1L)
)
with_second_set <- genefunnel(
    benchmark_profiles,
    list(
        pathway = rownames(benchmark_profiles),
        first_pair = rownames(benchmark_profiles)[1:2]
    ),
    BPPARAM = serial
)

stopifnot(
    identical(pattern_components$score, pattern_scores),
    identical(
        unname(pattern_scores["pathway", ]),
        c(8, 9, 26 / 3, 8 / 3, 0)
    ),
    identical(
        unname(pattern_components$observed_sum["pathway", ]),
        c(8, 9, 10, 8, 8)
    ),
    identical(
        unname(pattern_components$penalty["pathway", ]),
        c(0, 0, 4 / 3, 16 / 3, 8)
    ),
    identical(
        unname(pattern_components$balance["pathway", ]),
        c(1, 1, 13 / 15, 1 / 3, 0)
    ),
    identical(
        unname(pattern_components$effective_size["pathway", ]),
        rep(4L, 5L)
    ),
    identical(
        unname(pattern_scores["pathway", ]),
        unname(isolated_scores)
    ),
    identical(
        unname(pattern_scores["pathway", ]),
        unname(with_second_set["pathway", ])
    )
)
```

The coordinated shift changes magnitude without changing balance. Partial
activation increases the total but also incurs a penalty. The matched-total
comparison between `reference` and `same_sum_uneven` isolates the effect of
within-set absolute dispersion. `one_nonzero` retains the same total as the
reference but scores zero. Scoring every column separately and adding a second
gene set leave the pathway scores exactly identical, directly checking sample
and gene-set independence.

These are mechanism checks, not evidence that arithmetic balance is biological
co-expression, correlation, coordinated regulation, pathway truth, or
superiority over another method.

## Comparative and empirical designs

Before inspecting results, declare the benchmark's data-generating model,
biological unit, preprocessing, gene universe, and set catalogue. Also declare
the signal and null sets, downstream test, multiplicity rule, seed schedule,
and decision threshold. Pair methods on the same independently generated
replicates. Vary partial activation, mean-preserving dispersion changes, and
coordinated shifts separately because they test different mechanisms.

The thesis used a 20,000-gene matrix of negative-binomial counts. It included
1,000 disjoint 20-member sets, 50 signal sets, 950 null sets, and five samples
per group. It applied TMM normalization and compared
ORA, `camera`, and `fgsea` with GSVA, ssGSEA, and GeneFunnel score pipelines.
The latter three were tested with limma-trend; the first three used their own
inference. The study reported no independent simulation repetitions, complete
seed schedule, or Monte Carlo uncertainty. Therefore, its outcomes represent
one controlled study, not a general performance estimate.

Include direct sum/mean ablations and a no-change control before adding a large
competitor panel. ORA, competitive tests, ranked-list enrichment, and
functional class scoring test different hypotheses; a cross-family comparison
therefore evaluates end-to-end pipelines rather than scoring algorithms alone.
Within methods that share a hypothesis and input contract, use a common
compatible input to isolate scoring rules. Label any method-native preprocessing
analysis as a separate regime.

At each replicate, retain TP, FP, TN, and FN. Also retain sensitivity,
specificity, precision, F1, and the observed false discovery proportion (FDP).
If `R = TP + FP` sets are called, define `FDP = FP / max(R, 1)`, so no calls
give FDP zero. FDR is `E(FDP)`; mean FDP across independent replicates estimates
it and requires Monte Carlo uncertainty. Declare zero-denominator conventions
for precision and F1. Accuracy alone is weak when null sets greatly outnumber
signal sets. A mean BH-adjusted p-value among planted signal sets is not an FDR
and must be labelled as an adjusted-p-value summary among signals.

Sensitivity analysis should report a profile rather than only the most
favourable point. Useful axes include activated fraction, effect magnitude,
within-set dispersion, set size and overlap, member correlation, sample count,
zero density, missingness, assay coverage, and catalogue choice. Parameter
selection and final evaluation require separate realizations or a held-out
split.

When sets overlap, a nominal null sharing an edited member is affected by
construction. Use disjoint evaluation nulls or prospectively define which sets
become affected through shared members. Report that spillover separately rather
than counting it as a false positive.

When synthetic distributions are inadequate, make known changes against a real
data background. Start with a set that does not overlap the rest of the test
catalogue. Then repeat with an overlapping set to expose spillover from shared
edited members. The thesis set all target members to their pre-edit set mean. This
change removed within-set variation while preserving the total. The thesis
also added 100 counts to every target member in a separate test.

Preserve donor or other biological-unit labels. Condition-pooled profiles can
illustrate scoring. For inferential pseudobulk, sum assay counts within each
biological unit, condition, and cell-type stratum before scoring. Score every
unit separately. Treat units, rather than cells or technical repeats, as
replicates. Include unit/block effects only for paired or repeated designs.
Pooling biological units before scoring changes the nonlinear estimand. An
unmodified real-data contrast has no known pathway truth. Use it for
plausibility and inspect the member values that drive top-ranked sets.

## Reproducible resource evidence

The source repository contains an explicitly versioned
[protocol](https://github.com/eturkes/genefunnel/tree/main/benchmark).
The thesis resource study used five `bench::mark` iterations with six serial
pseudobulk samples, 60 parallel samples, and a halving series up to 600 cells.
Those coupled changes motivate current scaling scenarios but do not isolate
parallel speedup from workload size.

With `genefunnel` installed in a library visible to `Rscript --vanilla`, run
the controlled and smoke suites from the source root. The linked protocol shows
the alternative commands for its project-local library. The full command is
potentially expensive:

```{sh benchmark-protocol, eval=FALSE}
Rscript --vanilla \
    benchmark/run-protocol.R --protocol=1.0.0 --suite=controlled

Rscript --vanilla \
    benchmark/run-protocol.R --protocol=1.0.0 --suite=performance \
    --preset=smoke --repeats=1 --workers=2

# Full workload: 40 isolated observations at five repeats per scenario.
Rscript --vanilla \
    benchmark/run-protocol.R --protocol=1.0.0 --suite=performance \
    --preset=full --repeats=5 --workers=2
```

The full preset contains 60-sample dense bulk-like and 600-sample sparse
single-cell-like workloads; each crosses low/high set overlap and serial versus
two-worker SOCK execution. Storage, density, and sample count remain coupled
scenario classes rather than isolated effects. Fixtures and seeds are shared
across backends, and serialized output digests must agree.

Isolated measurement passes record elapsed time and, where Unix `/proc` is
available, 10-ms sampled aggregate resident memory for the scoring process
tree. Where R allocation profiling is available, a separate pass records
cumulative manager-side R allocations, not retained or native memory.
Unavailable fields remain `NA`. Historical thesis timings are
environment-specific rather than current guarantees, so benchmark the intended
workload.

Executable protocol 1.0.0 contains analytic controls and descriptive resource
workloads, but no competitors, external assays, or thesis-data reproduction.
The separate prospective comparative parent `validation-protocol.md` and
immutable effective-design amendment `validation-amendment.md` in that
benchmark directory are designs rather than results. The current package
therefore makes no general superiority or biological-validation claim.

# Identifiers and preprocessing

Matrix row names and gene-set names must be unique, non-missing, and non-empty.
Member matching is exact, with no case conversion, synonym resolution, or
annotation lookup. Duplicated members within one set are counted in coverage
diagnostics and deduplicated for scoring. Overlapping sets remain valid and do
not influence one another.

GeneFunnel requires non-negative input with a meaningful zero. It applies no
automatic preprocessing. An arbitrary positive shift is unsuitable because
adding a constant $c$ to all $n$ observed members adds $nc$ to the score. Raw
non-negative measurements can be preferable when scientifically appropriate,
but the package does not prescribe a universal pipeline position.

# Interpretation and limitations

The absolute-deviation penalty treats within-set variability as evidence
against coherent activity. That is a scientific assumption rather than a
universal property of biological pathways.

The sum term preserves magnitude and therefore also preserves effects of input
units, library size, preprocessing, set size, and feature coverage. Independent
sample scoring does not make scores from unrelated data sets automatically
comparable. Such comparisons require compatible units, preprocessing,
identifiers, gene-set definitions, versions, and coverage policies.

Gene-set database revisions can change scores. The package does not guarantee
that the resulting distribution meets the assumptions of every downstream
statistical method. Validate those assumptions for the intended analysis.

The complete normative contract is installed as `SCIENTIFIC_SPEC.md`:

```{r specification-path}
specification <- system.file("SCIENTIFIC_SPEC.md", package = "genefunnel")
stopifnot(nzchar(specification), file.exists(specification))
basename(specification)
```

# Citation

The package citation includes the software and the thesis that defines its
scientific intent.

```{r citation}
citation("genefunnel")
```

# Session information

```{r session-info}
sessionInfo()
```
