---
title: "GeneFunnel: scientific scoring and practical use"
author: "Emir Turkes"
date: "18 July 2026"
package: genefunnel
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
    math_method: mathml
vignette: >
  %\VignetteIndexEntry{GeneFunnel: scientific scoring and practical use}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Assisted-by: OpenAI Codex. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.align = "center"
)
Sys.setenv(TZ = "UTC")
```

# Introduction

GeneFunnel turns a non-negative feature-by-sample matrix into a
gene-set-by-sample score matrix. It preserves observed magnitude while
penalizing absolute variation among the available members of each set. Every
gene set and sample is calculated independently.

Transcriptomic and proteomic assays commonly use Bioconductor's matrix,
sparse-matrix, and parallel infrastructure. GeneFunnel fits that environment by
accepting base and `Matrix` matrices and a `BiocParallelParam`, while returning
a plain numeric matrix for downstream analysis. It is a scorer only: it does
not normalize data, impute missing values, map identifiers, retrieve gene sets,
or perform statistical testing.

The package is in development and has not been accepted into Bioconductor. Its
source and issue tracker are on the
[GeneFunnel repository](https://github.com/eturkes/genefunnel).

# Installation

Install the development version from GitHub with Bioconductor-compatible
dependencies:

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("eturkes/genefunnel")
```

Load the installed package and choose a serial backend for the first examples.

```{r load}
library(genefunnel)
serial <- BiocParallel::SerialParam()
```

# Score definition

For one declared gene set, GeneFunnel first deduplicates member identifiers and
matches them exactly and case-sensitively to matrix row names. Within sample
$j$, it omits matched values that are `NA` or `NaN`. If the remaining observed
values are $x_1,\ldots,x_n$ and $n \ge 2$, the score is

$$
\bar{x}=\frac{1}{n}\sum_{i=1}^{n}x_i
$$

and

$$
\operatorname{GeneFunnel}(x_1,\ldots,x_n)
=\sum_{i=1}^{n}x_i
-\frac{n}{2(n-1)}\sum_{i=1}^{n}|x_i-\bar{x}|.
$$

The cell is `NA_real_` when fewer than two observed members remain. Zero is an
observed value, including an implicit sparse zero, so it participates in $n$,
the sum, the mean, and the deviation. Infinities and negative values are
invalid.

The following matrix places the canonical cases in columns.

```{r canonical}
canonical <- matrix(
    c(4, 4, 4, 0, 0, 0, 4, 0, 0, 1, 2, NA, 4, NA, NA),
    nrow = 3,
    dimnames = list(
        c("A", "B", "C"),
        c("equal", "zeros", "one_nonzero", "missing_pair", "one_observed")
    )
)
canonical_sets <- list(
    three_members = c("A", "B", "C"),
    two_members = c("A", "B")
)

canonical_scores <- genefunnel(
    canonical,
    canonical_sets,
    BPPARAM = serial
)
knitr::kable(canonical_scores, digits = 3)

stopifnot(
    identical(
        unname(canonical_scores["three_members", ]),
        c(12, 0, 0, 2, NA_real_)
    ),
    identical(
        unname(canonical_scores["two_members", ]),
        c(8, 0, 0, 2, NA_real_)
    )
)
```

Equal values score their sum. All-zero sets and sets with only one non-zero
member score zero. The missing-pair column scores the observed values `c(1, 2)`
as 2; the last column has only one observed value and returns `NA`.

# Explain score components

`genefunnel_components()` calculates the same authoritative score while
exposing the arithmetic facts hidden by one scalar. The diagnostic native path
reads each input chunk once and returns aligned matrices.

```{r components}
canonical_components <- genefunnel_components(
    canonical,
    canonical_sets,
    BPPARAM = serial
)
component_table <- data.frame(
    sample = colnames(canonical),
    score = canonical_components$score["three_members", ],
    observed_sum = canonical_components$observed_sum["three_members", ],
    penalty = canonical_components$penalty["three_members", ],
    balance = canonical_components$balance["three_members", ],
    effective_size = canonical_components$effective_size["three_members", ],
    observed_fraction = canonical_components$observed_fraction[
        "three_members",
    ],
    row.names = NULL,
    check.names = FALSE
)
knitr::kable(component_table, digits = 3)

stopifnot(
    identical(canonical_components$score, canonical_scores),
    identical(
        unname(canonical_components$observed_sum["three_members", ]),
        c(12, 0, 4, 3, 4)
    ),
    is.na(canonical_components$balance["three_members", "zeros"]),
    identical(
        canonical_components$status$semantic[
            "three_members",
            "one_observed"
        ],
        "too_few_observed"
    )
)
```

For a positive total, `balance` is Bulla's evenness index on the observed
within-set shares and the exact factorization is `score = observed_sum *
balance`. It describes arithmetic evenness, not co-expression or pathway
truth. `observed_fraction` uses globally matched members as its denominator;
declared-set coverage remains a separate fact.

Finite sums, penalties, or balances can overflow or underflow an R double even
when the score is representable. `status` distinguishes ordinary, binary-scaled,
and unavailable diagnostics. A scaled value uses the corresponding `mantissa`
and `exponent` matrices under `canonical_components$scaled`; ordinary values
have `NA` there. The `conditioning` matrix marks cells where rounded component
reconstruction is unsafe.

# Coverage is caller controlled

A declared member absent from the matrix changes global assay coverage. It is
different from a missing cell, which affects only one sample, and from an
observed zero. `gene_set_coverage()` exposes these facts before scoring.

```{r coverage}
catalogue <- list(
    complete = c("A", "B"),
    partial = c("A", "B", "D"),
    duplicated = c("A", "A", "B"),
    insufficient = c("A", "Z")
)
coverage <- gene_set_coverage(catalogue, rownames(canonical))
knitr::kable(coverage, digits = 3)
```

GeneFunnel can score a set whenever at least two unique members match. It does
not impose a coverage fraction because assay-specific policies differ. A strict
transcriptomic analysis might require complete coverage:

```{r strict-coverage}
strict <- !is.na(coverage$coverage) & coverage$coverage == 1
names(catalogue[strict])
```

A low-coverage proteomic analysis might require at least 50% coverage and at
least two matched members:

```{r relaxed-coverage}
relaxed <- !is.na(coverage$coverage) &
    coverage$coverage >= 0.5 &
    coverage$matched_size >= 2
names(catalogue[relaxed])

relaxed_scores <- genefunnel(
    canonical,
    catalogue[relaxed],
    BPPARAM = serial
)
knitr::kable(relaxed_scores, digits = 3)
```

These thresholds are examples, not package recommendations. Record the exact
gene-set collection and version alongside the chosen policy.

# Dense and sparse matrices

Sparse input stays sparse through scoring, and implicit zeros remain observed.
Stored sparse `NA` and `NaN` values follow the same sample-specific omission
rule as dense missing values.

```{r sparse}
sparse_canonical <- Matrix::Matrix(canonical, sparse = TRUE)
sparse_scores <- genefunnel(
    sparse_canonical,
    canonical_sets,
    BPPARAM = serial
)

isTRUE(all.equal(
    sparse_scores,
    canonical_scores,
    tolerance = 0,
    check.attributes = TRUE
))
```

The dense and sparse implementations are required to produce the same named
matrix. Sparse storage changes memory use and traversal, not the scientific
calculation.

# Serial and parallel execution

`BPPARAM` accepts any suitable `BiocParallelParam`. `SerialParam()` is useful
for debugging and small inputs. A portable SOCK backend can distribute bounded
column chunks while preserving sample order.

```{r parallel}
sock <- BiocParallel::bpstart(BiocParallel::SnowParam(
    workers = 2L,
    type = "SOCK",
    progressbar = FALSE
))
parallel_scores <- tryCatch(
    genefunnel(canonical, canonical_sets, BPPARAM = sock),
    finally = BiocParallel::bpstop(sock)
)

identical(parallel_scores, canonical_scores)
```

Process startup and data transfer can outweigh parallel work on small inputs.
Benchmark the intended workload rather than assuming that more workers are
faster.

# Identifiers and preprocessing

Matrix row names and gene-set names must be unique, non-missing, and non-empty.
Member matching is exact, with no case conversion, synonym resolution, or
annotation lookup. Duplicated members within one set are counted in coverage
diagnostics and deduplicated for scoring. Overlapping sets remain valid and do
not influence one another.

GeneFunnel requires non-negative input with a meaningful zero. It applies no
automatic preprocessing. An arbitrary positive shift is unsuitable because
adding a constant $c$ to all $n$ observed members adds $nc$ to the score. Raw
non-negative measurements can be preferable when scientifically appropriate,
but the package does not prescribe a universal pipeline position.

# Interpretation and limitations

The absolute-deviation penalty treats within-set variability as evidence
against coherent activity. That is a scientific assumption rather than a
universal property of biological pathways.

The sum term preserves magnitude and therefore also preserves effects of input
units, library size, preprocessing, set size, and feature coverage. Independent
sample scoring does not make scores from unrelated data sets automatically
comparable. Such comparisons require compatible units, preprocessing,
identifiers, gene-set definitions, versions, and coverage policies.

Gene-set database revisions can change scores. The package does not guarantee
that the resulting distribution meets the assumptions of every downstream
statistical method. Validate those assumptions for the intended analysis.

The complete normative contract is installed as `SCIENTIFIC_SPEC.md`:

```{r specification-path}
specification <- system.file("SCIENTIFIC_SPEC.md", package = "genefunnel")
stopifnot(nzchar(specification), file.exists(specification))
basename(specification)
```

# Citation

The package citation includes the software and the thesis that defines its
scientific intent.

```{r citation}
citation("genefunnel")
```

# Session information

```{r session-info}
sessionInfo()
```
